---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastadi

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![CRAN status](https://www.r-pkg.org/badges/version/fastadi)](https://CRAN.R-project.org/package=fastadi)
[![Codecov test coverage](https://codecov.io/gh/RoheLab/fastadi/branch/master/graph/badge.svg)](https://codecov.io/gh/RoheLab/fastadi?branch=master)
[![R build status](https://github.com/RoheLab/fastadi/workflows/R-CMD-check/badge.svg)](https://github.com/RoheLab/fastadi/actions)
<!-- badges: end -->

`fastadi` implements the `AdaptiveImpute` matrix completion algorithm. `fastadi` is a self-tuning alternative to algorithms such as `SoftImpute` (implemented in the [`softImpute`](https://cran.r-project.org/package=softImpute) package), truncated SVD, maximum margin matrix factorization, and weighted regularized matrix factorization (implemented in the [`rsparse`](https://github.com/rexyai/rsparse) package). In simulations `fastadi` often outperforms `softImpute` by a small margin.

You may find `fastadi` useful if you are developing embeddings for sparsely observed data, if you are working in natural language processing, or building a recommendation system.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("RoheLab/fastadi")
```

## Example usage

Here we embed users and items in the MovieLens 100K dataset.

```{r}
library(fastadi)

mf <- adaptive_impute(ml100k, rank = 3L, max_iter = 5L)
```

```{r}
mf
```

Note that the vignettes are currently scratch work for reference by the developers and are not yet ready for general consumption.

## References

1. Cho, J., Kim, D. & Rohe, K. Asymptotic Theory for Estimating the Singular Vectors and Values of a Partially-observed Low Rank Matrix with Noise. arXiv:1508.05431 [stat] (2015). http://arxiv.org/abs/1508.05431

2. Cho, J., Kim, D. & Rohe, K. Intelligent Initialization and Adaptive Thresholding for Iterative Matrix Completion; Some Statistical and Algorithmic Theory for Adaptive-Impute. Journal of Computational and Graphical Statistics 1–26 (2018) doi:10.1080/10618600.2018.1518238. https://amstat.tandfonline.com/doi/abs/10.1080/10618600.2018.1518238

You can find the original implementation accompanying these papers [here](https://github.com/chojuhee/hello-world).

